---
title: 'Data Visualisation: Telling Visual Stories From Data'
author: "Bevan Koopman"
date: "11 /04/2017"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(reshape2)
theme_set(theme_bw())
```


## Anscombe's quartet

```{r}
anscombe_m <- read.csv("datasets/anscombe_m.csv")
summary(anscombe_m)
```

Checks the mean, variance for each dataset

```{r}
aggregate(y ~ Dataset, anscombe_m, mean)
aggregate(x ~ Dataset, anscombe_m, mean)
aggregate(y ~ Dataset, anscombe_m, var)
aggregate(x ~ Dataset, anscombe_m, var)
for(d in c("A", "B", "C", "D")) {
  print(cor(anscombe_m[anscombe_m$Dataset==d,]$x, anscombe_m[anscombe_m$Dataset==d,]$y))
}
```

Now plot the four different data sets
```{r}
ggplot(anscombe_m, aes(x=x, y=y)) + geom_point(size=3)  + facet_wrap(~Dataset) 
```

## Grammar of graphical elements

### 1. The `Data` layer

Have a look at the `diamonds` dataset.
```{r}
summary(diamonds)
```



### 2. The `Aesthetics` layer

Now we map parts of the data to some graphical elements (x, y, color, size, etc)
```{r}
ggplot(diamonds, aes(x=carat, y=price)) 
```

Note, that there is no notion at the moment of the types of plot - that is the next layer!

### 3. The `Geometries` layer

Now we define a geometry (this case, a scatter geometry):
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point(alpha=.4)
```

### 4. The `Facet` layer

Now lets split the plot up by `clarity`:
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_jitter(alpha=.4) + facet_wrap(~clarity)
```

### 5. The `Statistics` layer

Now let's add an graphical element that aids in understanding, namely a linear model:
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_jitter(alpha=.4) + facet_wrap(~clarity) + geom_smooth(method="lm", se=F)
```


### 6. The `Coordinates` layer

Now let's ajust the scales and axis levels. Specifically, we'll restrict x-axis carat to be between 0 and 3, report x-axis carat in 0.5 increments, change y-axis price to be in dollar format and provide x and y axis labels.
```{r}
library(scales) # used to print things in dollar ($) format
ggplot(diamonds, aes(x=carat, y=price)) + geom_jitter(alpha=.4) + facet_wrap(~clarity) + geom_smooth(method="lm", se=F) + scale_x_continuous(limits=c(0,3), breaks = seq(0,3, 0.5)) + scale_y_continuous(labels = dollar) + xlab("Mass (carat)") + ylab("Price (US$)")
```

### 6. The `Theme` layer

Now just change the look and feel:
```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_jitter(alpha=.4) + facet_wrap(~clarity) + geom_smooth(method="lm", se=F) + scale_x_continuous(limits=c(0,3), breaks = seq(0,3, 0.5)) + scale_y_continuous(labels = dollar) + xlab("Mass (carat)") + ylab("Price (US$)") + theme_minimal()
```


## Data that supports Visualisation

Have a look at the `airquality` dataset.

```{r}
summary(airquality)
```

Get the `reshape2` package (not `reshape`).
```{r}
require(reshape2)
```

Look at the top 10 rows of the data:
```{r}
head(airquality)
```


Transform to long format.
```{r, results='hide'}
head(melt(airquality))
```

Tranform to long format but keep `Month` and `Day` as columns.
```{r}
long_airquality <- melt(airquality, id.vars=c('Month', 'Day'), variable.name = "Measurement")
head(long_airquality)
```

Now the data is in a easier format for plotting.
```{r}
ggplot(long_airquality, aes(x=Month, y=value)) + geom_point() + facet_wrap(~Measurement, scales="free")
```

Calculate a montly average
```{r}
agg_long_airquality <- aggregate(value ~ Month + Measurement, long_airquality, mean)
ggplot(agg_long_airquality, aes(x=Month, y=value)) + geom_point() + geom_line() + facet_wrap(~Measurement, scales="free")
```

But rather than calculate the average, why not just use a boxplot:
```{r}
ggplot(long_airquality, aes(x=Month, y=value)) + geom_jitter() + facet_wrap(~Measurement, scales="free") # the old way
ggplot(long_airquality, aes(x=Month, y=value)) + geom_jitter() + facet_wrap(~Measurement, scales="free") + stat_summary(fun.y="mean", color="red", size=3, geom="point")
ggplot(long_airquality, aes(x=factor(Month), y=value)) + geom_boxplot(varwidth = T) + facet_wrap(~Measurement, scales="free") + scale_x_discrete()
```


```{r}

```

## Different Plot Types

Let's return to `diamonds` dataset and looks through a bunch of different plot types and what they show.

Just a reminder of the diamonds dataset:
```{r}
summary(diamonds)
```


### The scatter plot

```{r}
ggplot(diamonds, aes(x=carat, y=price)) + geom_point()
```

Now let's add a visual distinction (as color) for diamond `clarity`:
```{r}
ggplot(diamonds, aes(x=carat, y=price, color=clarity)) + geom_point()
```

Now lets add another categorical dimension for the quality of the `cut` as the size:
```{r}
ggplot(diamonds, aes(x=carat, y=price, color=clarity, size=cut)) + geom_point()
```

### Trends

Now let's apply a smoothing trend:
```{r}
ggplot(diamonds, aes(x=carat, y=price, color=clarity)) + geom_point() + geom_smooth(se=F)
ggplot(diamonds, aes(x=carat, y=price, color=clarity)) + geom_smooth(se=F)
```

### Faceting

The plot is pretty busy so lets rather divide things up according to the categorical data.

First divide by cut:
```{r}
ggplot(diamonds, aes(x=carat, y=price, color=clarity)) + geom_point() + facet_wrap(~cut)
```

Maybe by clarity instead:
```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + geom_point() + facet_wrap(~clarity)
```

And add a further facet for color:
```{r}
ggplot(diamonds, aes(x=carat, y=price, color=cut)) + geom_point() + facet_grid(color~clarity) + scale_color_grey()
```

### Barplot

Barplot are used for make comparisons. Let's have a look at some climate data.
```{r}
global_temps <- read.table("647_Global_Temperature_Data_File.txt", col.names = c("Year", "TempChange", "TempChange5Year"))
```

The data shows change in global surface temperature for a 136-year period, relative to 1951-1980 average temperatures.

```{r}
ggplot(global_temps, aes(x=Year, y=TempChange)) + geom_bar(stat="identity")
```

Let's add the bounds for the comparitive years:
```{r}
ggplot(global_temps, aes(x=Year, y=TempChange)) + geom_bar(stat="identity") + geom_vline(xintercept=c(1951,1980), color="red")
```


### Line plots

Line plots are designed to show trends. 

Let's look at the change in temps over the years:
```{r}
ggplot(global_temps, aes(x=Year, y=TempChange)) + geom_line()
```

Sometimes we want a trend and to compare indvidual years - what should we do?:
```{r}
ggplot(global_temps, aes(x=Year, y=TempChange)) + geom_line() + geom_point()
```

And adding the five-year average:
```{r}
ggplot(global_temps, aes(x=Year, y=TempChange)) + geom_line() + geom_point() + geom_line(aes(y=TempChange5Year, color="red"))
```

But this is not nice - better to format your data to support vis - i.e., a long format:
```{r}
long_global_temps = melt(global_temps, id.vars = "Year", variable.name = "MeasurementType", value.name = "Degrees")
```

Now we can go a better job:
```{r}
ggplot(long_global_temps, aes(x=Year, y=Degrees, color=MeasurementType)) + geom_line() + geom_point(data=subset(long_global_temps, MeasurementType=="TempChange"))
```

### Histograms

Histograms are used to convey an understanding about the distribution of your data. For example, the distribution on `price`:

```{r}
ggplot(diamonds, aes(x=price)) + geom_histogram()
```

```{r}
ggplot(diamonds, aes(x=price, fill=cut)) + geom_histogram()
```


### Boxplots


Box plots provide a way of showing distributions and averages. Let's return to the `airquality` dataset:
```{r}
ggplot(long_airquality, aes(x=Month, y=value)) + geom_jitter() + facet_wrap(~Measurement, scales="free") # the old way
ggplot(long_airquality, aes(x=factor(Month), y=value, fill=Measurement)) + geom_boxplot(varwidth = T) + facet_wrap(~Measurement, scales="free") # the old way
```

What us going on with Ozone measurements for June:
```{r}
subset(long_airquality, Measurement=="Ozone" & Month==6)
```

### Heatmap

A heat map is a two-dimensional representation of data in which values are represented by colors. A simple heat map provides an immediate visual summary of information. More elaborate heat maps allow the viewer to understand complex data sets.

Let's use the diamond dataset to look at how clarity, color and price are related.

First we will do some aggregation on the data:
```{r}
agg_diamonds <- aggregate(price ~ clarity + color, diamonds, mean)
```


```{r}
ggplot(agg_diamonds, aes(x=clarity, y=color, fill=price)) + geom_tile(colour = "white") + scale_fill_gradient(low = "white", high = "steelblue")
```

